Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-08-11 manuscript/manuscript.Rmd
minor updates.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

@@

The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

>>>>

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].

@@

The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
